---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo = TRUE}
data <- read.csv("activity.csv", header = TRUE, sep = ",")
clean_data <- na.omit(data)
```


## What is mean total number of steps taken per day?


```{r, echo = TRUE}
#aggregate the data
total_steps_per_day <- aggregate(steps ~ date, data = clean_data, sum)
#create the histogram 
barplot(height = total_steps_per_day$steps, 
        names.arg = total_steps_per_day$date, 
        main = "Total Steps Taken Each Day", 
        xlab = "Date", 
        ylab = "Total Steps", 
        col = "blue", 
        las = 2, 
        cex.names = 0.7) 

#the the mean and median of the total number of steps taken per day

mean_steps <- mean(total_steps_per_day$steps)
median_steps <- median(total_steps_per_day$steps)

```

The mean of the total number of steps taken per day is `r mean_steps`

The median of the total number of steps taken per day is `r median_steps`

## What is the average daily activity pattern?

```{r, echo=TRUE}
#necessary package 
library(dplyr)

#aggregate the data 
average_steps_per_interval <- aggregate(steps ~ interval, data = clean_data, mean)
with(average_steps_per_interval, plot(interval, steps, type="l", xlab = "5-minute Interval", ylab = "Average Number of Steps", main = "Average Number of Steps per 5-minute Interval"))
#the interval with the maximun average number of steps 
max_interval <- average_steps_per_interval %>%
  filter(steps == max(steps))
max_interval
```
The interval with the maximum average steps is `r max_interval`

## Imputing missing values

```{r, echo=TRUE}
#The total number of Na 
total_na <- sum(is.na(data))
total_na
# The mean steps for each interval
mean_steps_per_interval <- data %>%
  group_by(interval) %>%
  summarize(mean_steps = mean(steps, na.rm = TRUE))

mean_steps_per_interval
# Merge mean steps with the original data
data <- merge(data, mean_steps_per_interval, by = "interval", all.x = TRUE)
# Replace NA values in the steps column with corresponding mean steps
data$steps <- ifelse(is.na(data$steps), data$mean_steps, data$steps)

# Clean up by removing the mean_steps column
data <- data[, !(names(data) %in% "mean_steps")]
```
```{r, echo=TRUE}
##create the histogram of the total number of steps taken each day using data with no Na
##aggregate the data
total_steps_per_day1 <- aggregate(steps ~ date, data = data, sum)

#the histogram 
barplot(height = total_steps_per_day1$steps, 
        names.arg = total_steps_per_day1$date, 
        main = "Total Steps Taken Each Day", 
        xlab = "Date", 
        ylab = "Total Steps", 
        col = "orange", 
        las = 2, 
        cex.names = 0.7) 

#the the mean and median of the total number of steps taken per day

mean_steps1 <- mean(total_steps_per_day1$steps)
median_steps1 <- median(total_steps_per_day1$steps)

```
The mean of the total number of steps taken per day when the Nas are taken into consideration is `r mean_steps1`

The median of the total number of steps taken per day when the Nas are taken into consideration is `r median_steps1`

the values of the mean and median does not differ between the two data sets. Imputing missing data did not affect the estimation of the total daily number of steps

## Are there differences in activity patterns between weekdays and weekends?


```{r, echo=TRUE}
#convert the date column to date format
data$date <- as.Date(data$date, format = "%Y-%m-%d")
#create a new factor variable that determine if the day of the week is weekend or not 
data$day_type <- ifelse(weekdays(data$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
data$day_type <- as.factor(data$day_type)

# Print the first few rows to verify the new variable
head(data)


```


```{r, echo=TRUE}
##create the plot 
# Subset the data
data_weekday <- subset(data, day_type == "weekday")
data_weekend <- subset(data, day_type == "weekend")
# Calculate the average steps for each interval for weekdays
average_steps_weekday <- aggregate(steps ~ interval, data = data_weekday, mean)

# Calculate the average steps for each interval for weekends
average_steps_weekend <- aggregate(steps ~ interval, data = data_weekend, mean)

# Determine the common y-axis limits
ylim <- range(c(average_steps_weekday$steps, average_steps_weekend$steps))

# Set up the plotting area to have two plots, one on top of the other
par(mfrow = c(2, 1), oma = c(4, 4, 2, 1), mar = c(1, 1, 2, 1))

# Plot for weekdays
with(average_steps_weekday, plot(interval, steps, type = "l", ylim = ylim, xaxt = "n", ylab = "", main = "Weekday"))
mtext("Average Number of Steps", side = 2, line = 3, outer = TRUE)

# Plot for weekends
with(average_steps_weekend, plot(interval, steps, type = "l", ylim = ylim, ylab = "", main = "Weekend"))
mtext("5-minute Interval", side = 1, line = 2, outer = TRUE)

```

The person is more active in weekdays compared to weekends 



